\pkg{sparseHessianFD}: An R Package for Estimating Sparse Hessian Matrices

### 2016-12-25 nobuild/jss_revision/sparseHessianFD.Rnw
proof of sparsity with complex step method

However, for many problems, the Hessian is
\emph{sparse}, meaning that the proportion of  ``structural''  zeros
(Hessian matrix elements that are always zero, regardless of the value at which
function is estimated) is high.  Consider a log
posterior density in a Bayesian hierarchical
model.  If the outcomes across units are conditionally
independent, the cross-partial derivatives of heterogeneous variables
across units are zero.  As the number of units
increases, the size of the Hessian still grows quadratically, but the number
of \emph{non-zero} elements grows only linearly; the Hessian
becomes increasingly sparse.  The row and column indices of the
non-zero elements comprise the \emph{sparsity pattern} of the Hessian,
and are typically known in advance, before computing the values of
those elements.  \proglang{R} packages such as \pkg{trustOptim}
\citep{R_trustOptim}, \pkg{sparseMVN} \citep{R_sparseMVN} and
\pkg{ipoptr} \citep{R_ipoptr} have the capability to accept Hessians
in a compressed sparse format.

The \pkg{sparseHessianFD} package is a tool for estimating
sparse Hessians numerically, using either finite differences or
complex perturbations of
gradients.  Section~\ref{sec:numdiff} will
cover the specifics, but the basic idea is as follows.  Consider a real-valued
function $f(x)$, its gradient $\nabla{f(x)}$, and its Hessian
$\hess{}{f(x)}$, for $x\in\mathbb{R}^M$.  Define the derivative vector
as the transpose of the gradient, and a vector of partial derivatives, so
${\parD{}{}{f(x)}=\nabla f(x)^\top=\left(\parD{}{1}{},\mathellipsis,\parD{}{M}{}\right)}$.   (Throughout the paper, we will try
to reduce notational clutter by referring to the derivative and
Hessian as $\parD{}{}{}$ and $\hess{}{}$, respectively, without the
$f(x)$ symbol).  Let $e_m$ be a vector of zeros, except with a 1 in the $m$th
element, and let $\delta$ be a sufficiently small scalar
constant. A ``finite difference'' linear approximation to the $m$th column of the Hessian is ${\hess{m}{}\approx\left(\nabla f(x+\delta
  e_m)- \nabla f(x)\right)/\delta}$.  Estimating a dense Hessian in this way involves at least $M+1$ calculations of the
gradient: one for the gradient at $x$, and one after perturbing each
of the $M$ elements of $x$, one at a time. If the function is complex
analytic, an alternative approximation that requires $M$ gradient
evaulations is $\hess{m}{}\approx\Imag\left(\nabla f(x+i\delta
  e_m)\right)/\delta$, where $i=\sqrt{-1}$ and $\Imag$ returns the imaginary part of a complex
number \citep{SquireTrapp1998}. This ``complex step'' method can be more accurate than the
finite difference method because there is no subtraction operation
that might generate roundoff errors, and thus $\delta$ can be made
arbitrarily small to avoid truncaton errors \citep{LaiCrassidis2008}. However, it does require that the
implementation of $f(x)$ accept a complex argument, and return a
complex result, even though we are interested in differentiating $f(x)$ only along
the real line.

Regardless of the approximation method used, if the Hessian is sparse, most of the
elements are constrained to zero.  Depending
on the sparsity pattern of the Hessian, those constraints may let us recover
the Hessian with fewer gradient evaluations by perturbing multiple
elements of $x$ together.  For some sparsity patterns, estimating a Hessian in
this way can be profoundly efficient.  In fact, for the hierarchical models that we consider in
this paper, the number of gradient evaluations does not increase with
the number of additional heterogeneous units.

@@

The partial derivative of a real scalar-valued function $f(x)$ with respect to $x_j$ (the $j$th
component of $x$) is defined as
\begin{align}
  \label{eq:defParD}
\parD{}{j}{f(x)}=\lim\limits_{\delta\to 0}\frac{f(x+\delta e_j)-f(x)}{\delta}
\end{align}
For a sufficiently small $\delta$, this definition allows for a
linear approximation to $\parD{}{j}{f(x)}$.  The derivative of $f(x)$
is the vector of all $M$ partial derivatives.
\begin{align}
  \parD{}{}{f(x)}=\left(\parD{}{1}{f(x)},\mathellipsis,\parD{}{M}{f(x)}\right)
  \end{align}
 The \emph{gradient} is defined as $\nabla f(x)=\parD{}{}{f(x)}^\top$.

@@

We are interested in estimating the Hessian only for real
$x\in\Real{M}$. However, when $f(x)$ is complex-valued over a complex domain $x\in\Complex{M}$, we can approximate $\parD{}{}{f(x)}$
and $\hess{}{f(x)}$ for real $x$ using the complex step
method.  This method comes from a Taylor
series expansion of $f(x)$ in the imaginary direction of the complex
plane \citep{SquireTrapp1998}.  After rearranging terms, and taking the imaginary
parts of both sides,
\begin{align}
  \label{eq:complexTaylor}
  f(x+i\delta)&=f(x) + i\delta\parD{}{}{f(x)}+\mathcal{O}(\delta^2)\\
  \parD{}{}{f(x)}&\approx\frac{\Imag(f(x+i\delta))}{\delta}
\end{align}

Estimating a first derivative using the complex step method does not
require a differencing operation, so there is no loss of precision
from rounding error, and the approximation of the derivative can be
made arbitrarily precise as $\delta\rightarrow 0$.  This is not the
case for second-order approximations of
the Hessian \citep{AbreuStich2013}.  However, when the gradient can be
computed exactly, we can compute a first-order approximation to
Hessian by treating it as the Jacobian of
a vector-valued function \citep{LaiCrassidis2008}.

@@

If this matrix were dense, we would need two evaluations of the
complex $\parD{}{}f(x)$ to estimate it.  But what if it had the same sparsity
pattern as the Hessian in Equation~\ref{eq:FDhess2sp}? First, we
assume that structural zeros remain zero for all
complex $x\in\Complex{M}$.  Next, subtract
$\Imag(\parD{}{}{f(x_1,x_2)})$ from each column of
$\hess{}{f(x)}$. When $x$ is real, the imaginary part of the gradient
is zero, so this operation has no effect on the value of the
Hessian. But the sparsity constraints ensure that
the following identities hold:

@@

Only one evaluation of the gradient is required.

@@

The evaluation of the Hessian is a finite differenced approximation,
so it is very close to, but not identical to, the true value, in terms
of mean relative difference.

@@

A call to the \class{sparseHessianFD} \method{hessian} method will
apply the complex step method if the 
\funcarg{complex=TRUE} was set in the initializer. The functions
passed as \funcarg{fn} and \funcarg{gr} must both
accept a complex argument, and return a complex
result, even though we are differentiating a real-valued function. Although \pkg{base} \proglang{R} supports complex arguments
for most basic mathematical functions, many common functions (e.g.,  \func{gamma}, \func{log1p},
\func{expm1}, and the probability distribution functions) do not have
complex implementations. Furthermore, the complex step method is valid
only if the function is holomorphic
(complex differentiable). The methods in \class{sparseHessianFD} do
\emph{not} check that this is the case for the function at hand. We convey the
following warning from the documentation of the \pkg{numDeriv} package
\citep{R_numDeriv}, which also implements the complex step method:
``Avoid this method if you do not know that your function is
suitable. Your mistake may not be caught and the results will be
spurious.''

@@

In short, the complex step method can be much more accurate than
finite differencing, but it comes with theoretical and implementation
restrictions that may limit its universality.

>>>>

However, for many problems, the Hessian is
\emph{sparse}, meaning that the proportion of  ``structural''  zeros
(matrix elements that are always zero, regardless of the value at which
function is estimated) is high.  Consider a log
posterior density in a Bayesian hierarchical
model.  If the outcomes across units are conditionally
independent, the cross-partial derivatives of heterogeneous variables
across units are zero.  As the number of units
increases, the size of the Hessian still grows quadratically, but the number
of \emph{non-zero} elements grows only linearly; the Hessian
becomes increasingly sparse.  The row and column indices of the
non-zero elements comprise the \emph{sparsity pattern} of the Hessian,
and are typically known in advance, before computing the values of
those elements.  \proglang{R} packages such as \pkg{trustOptim}
\citep{R_trustOptim}, \pkg{sparseMVN} \citep{R_sparseMVN} and
\pkg{ipoptr} \citep{R_ipoptr} have the capability to accept Hessians
in a compressed sparse format.

The \pkg{sparseHessianFD} package is a tool for estimating
sparse Hessians numerically, using either finite differences or
complex perturbations of
gradients.  Section~\ref{sec:numdiff} will
cover the specifics, but the basic idea is as follows.  Consider a real-valued
function $f(x)$, its gradient $\nabla{f(x)}$, and its Hessian
$\hess{}{f(x)}$, for $x\in\mathbb{R}^M$.  Define the derivative vector
as the transpose of the gradient, and a vector of partial derivatives, so
${\parD{}{}{f(x)}=\nabla f(x)^\top=\left(\parD{}{1}{},\mathellipsis,\parD{}{M}{}\right)}$.   (Throughout the paper, we will try
to reduce notational clutter by referring to the derivative and
Hessian as $\parD{}{}{}$ and $\hess{}{}$, respectively, without the
$f(x)$ symbol).  Let $e_m$ be a vector of zeros, except with a 1 in the $m$th
element, and let $\delta$ be a sufficiently small scalar
constant. A ``finite difference'' linear approximation to the $m$th column of the Hessian is ${\hess{m}{}\approx\left(\nabla f(x+\delta
  e_m)- \nabla f(x)\right)/\delta}$.  Estimating a dense Hessian in this way involves at least $M+1$ calculations of the
gradient: one for the gradient at $x$, and one after perturbing each
of the $M$ elements of $x$, one at a time. Under certain conditions, a
more accurate approximation is the ``complex step'' method:  $\hess{m}{}\approx\Imag\left(\nabla f(x+i\delta
  e_m)\right)/\delta$, where $i=\sqrt{-1}$ and $\Imag$ returns the
imaginary part of a complex
number \citep{SquireTrapp1998}.  Regardless of the approximation method used, if the Hessian is sparse, most of the
elements are constrained to zero.  Depending
on the sparsity pattern of the Hessian, those constraints may let us recover
the Hessian with fewer gradient evaluations by perturbing multiple
elements of $x$ together.  For some sparsity patterns, estimating a Hessian in
this way can be profoundly efficient.  In fact, for the hierarchical models that we consider in
this paper, the number of gradient evaluations does not increase with
the number of additional heterogeneous units.

@@

The partial derivative of a real scalar-valued function $f(x)$ with
respect to $x_j$ (the $j$th
component of $x\in\Real{M}$) is defined as
\begin{align}
  \label{eq:defParD}
\parD{}{j}{f(x)}=\lim\limits_{\delta\to 0}\frac{f(x+\delta e_j)-f(x)}{\delta}
\end{align}
For a sufficiently small $\delta$, this definition allows for a
linear approximation to $\parD{}{j}{f(x)}$.  The derivative of $f(x)$
is the vector of all $M$ partial derivatives.
\begin{align}
  \parD{}{}{f(x)}=\left(\parD{}{1}{f(x)},\mathellipsis,\parD{}{M}{f(x)}\right)
  \end{align}
 The \emph{gradient} is defined as $\nabla f(x)=\parD{}{}{f(x)}^\top$.

@@

If $f(x)$ is defined over a complex domain and is holomorphic, then we
can we can approximate $\parD{}{}{f(x)}$
and $\hess{}{f(x)}$ at real values of $x$ using the complex step
method. This method comes from a Taylor
series expansion of $f(x)$ in the imaginary direction of the complex
plane \citep{SquireTrapp1998}.  After rearranging terms, and taking the imaginary
parts of both sides,
\begin{align}
  \label{eq:complexTaylor}
  f(x+i\delta)&=f(x) + i\delta\parD{}{}{f(x)}+\mathcal{O}(\delta^2)\\
  \parD{}{}{f(x)}&\approx\frac{\Imag(f(x+i\delta))}{\delta}
\end{align}

Estimating a first derivative using the complex step method does not
require a differencing operation, so there is no subtraction operation
that might generate roundoff errors.  Thus, the approximation can be
made arbitrarily precise as $\delta\rightarrow 0$ \citep{LaiCrassidis2008}.  This is not the
case for second-order approximations of
the Hessian \citep{AbreuStich2013}.  However, when the gradient can be
computed exactly, we can compute a first-order approximation to
Hessian by treating it as the Jacobian of
a vector-valued function \citep{LaiCrassidis2008}.

@@

If this matrix were dense, we would need two evaluations of the
$\parD{}{}f(x)$ to estimate it.  If the matrix were sparse, with the same sparsity
pattern as the Hessian in Equation~\ref{eq:FDhess2sp}, and we
assume that structural zeros remain zero for all
complex $x\in\Complex{M}$, then we need only one evaluation.  Suppose
we were to subtract $\Imag(\parD{}{}{f(x_1,x_2)})$ from each column of
$\hess{}{f(x)}$. When $x$ is real, the imaginary part of the gradient
is zero, so this operation has no effect on the value of the
Hessian. But the sparsity constraints ensure that
the following identities hold all complex $x$.

@@

for real $x$.  Only one evaluation of the gradient is required.

@@

The default choice of method is \funcarg{complex=FALSE}, so the
evaluation of the Hessian is a finite differenced approximation,
so it is very close to, but not identical to, the true value, in terms
of mean relative difference.

@@

If \funcarg{complex=TRUE} in the initializer, the call to the \method{hessian} method will
apply the complex step method. To use this method, the functions
passed as \funcarg{fn} and \funcarg{gr} must both
accept a complex argument, and return a complex
result, even though we are differentiating a real-valued function. Although \pkg{base} \proglang{R} supports complex arguments
for most basic mathematical functions, many common functions (e.g.,  \func{gamma}, \func{log1p},
\func{expm1}, and the probability distribution functions) do not have
complex implementations. Furthermore, the complex step method is valid
only if the function is holomorphic. The methods in \class{sparseHessianFD} do
\emph{not} check that this is the case for the function at hand. We convey the
following warning from the documentation of the \pkg{numDeriv} package
\citep{R_numDeriv}, which also implements the complex step method:
``avoid this method if you do not know that your function is
suitable. Your mistake may not be caught and the results will be
spurious.''

@@

In short, the complex step method can be more accurate than
finite differencing, but it comes with theoretical and implementation
restrictions that may limit its universality.
